The truncated KdV framework for modeling anomalous waves induced by abrupt depth changes

### 2019-08-06 paper2_JNLS/paper.tex
NM

%----------------------------------------------------------%
% Theory
%----------------------------------------------------------%
\section{The truncated KdV framework}

@@

%----------------------------------------------------------------------------%
% Nondimensionalization
%----------------------------------------------------------------------------%
\subsection{Nondimensionalization of variable-depth KdV}

@@

The evolution of any functional $\Fcnl[u]$ is given by 
\begin{align}
\partial_t \Fcnl = \{ \Fcnl, \Ham \} = \int_{-\pi}^{\pi} \vard{F}{u} \sympJ \vard{\Ham}{u} \dx
\end{align}
where again $\Ham = \Hup$ if $t<0$ and $\Ham = \Hdn$ if $t>0$.
%----------------------------------------------------------------------------%
% Truncation
%----------------------------------------------------------------------------%
\subsection{The truncated KdV framework}
\label{tKdVSec}

Consider the Fourier series of the state variable
\begin{align}
&u(x,t) = \sum_{k=-\infty}^{\infty} \uhat_k e^{i k x} \, , \\
\label{uhat}
&\uhat_k = \frac{1}{2 \pi} \int_{-\pi}^{\pi} u(x) e^{-i k x} \dx
\end{align}
Since $u$ is real valued, we have $\uhat_{-k} = \uhat_{k}^*$ and since momentum vanishes $\uhat_0 = 0$.
Next, consider the Galerkin truncation at wave number $\Lambda$
\begin{align}
\uL(x,t) = \Proj u =
\sum_{\abs{k} \le \Lambda} \uhat_k e^{i k x} \, , \qquad
\end{align}
where $\Proj$ is a projection operator and \eqref{uhat} still holds. Projecting the KdV equation \eqref{dimlessKdV} onto the finite dimensional space gives the truncated KdV equation (TKdV)
\begin{align}
\label{TKdV}
&\pd{\uL}{t} +  \frac{1}{2} C_3 \drat^{-3/2} \,\Proj (\uL)^2 + C_2 \drat^{1/2} \, \frac{\partial^3 \uL}{\partial x^3} = 0
\qquad \text{for } x \in [-\pi,\pi] \\
&C_3 = \frac{3}{2} \pi^{3/2} \epsup \delup^{-1} \, , \quad
C_2 = \frac{\pi^3 \delup}{6 \lamfac^2} 
\end{align}
Note the additional projection operator in front of the quadratic term $\uL^2$, which removes the aliased modes of wavenumber larger than $\Lambda$. The constants are the same as before and have been repeated here for convenience.

Remarkably, the TKdV system \eqref{TKdV} retains nearly the same Hamiltonian structure described in Section \ref{HamiltonianSection}, with the only modification being the inclusion of the projection operator \cite{bajars2013weakly, majda2019statistical}. The piecewise defined Hamiltonian for TKdV is given by
\begin{equation}
\label{TruncHamiltonian}
\HLupdn = C_2 \dratupdn^{1/2} \, \Htwo[\uL] - C_3 \dratupdn^{-3/2} \, \Hthree[\uL]
\end{equation}
where $\Hthree$ and $\Htwo$ the defined exactly as before \eqref{H3H2}, but now are simply applied to the projected variable $\uL = \Proj u$.
Then TKdV \eqref{TKdV} can be expressed as
\begin{align}
\partial_t {\uL} = \partial_x \Proj \, \vard{\HLupdn}{\uL}
\end{align}
where the new symplectic operator is $\SympL = \partial_x \Proj$.
% Note: Bajars 2013 has some important details on the Hamiltonian on the truncated system.

The momentum and energy defined in \eqref{MomEn} remain invariants of TKdV, with the same normalization values $\Mo[\uL] = 0$ and $\En[\uL] = 1$. Note that Parseval's identity gives
\begin{equation}
\En[\uL] = 2 \pi \sum_{k=1}^{\Lambda} \abs{\uhat_k}^2 = 1
\end{equation}
% The above agrees with the first line on top of p. 4 in Majda and Qi J Stat Phys 2019.

The KdV system (i.e. $\Lambda = \infty$) possesses the Hamiltonian \eqref{Hamiltonian} and momentum and energy \eqref{MomEn} and an infinite number of additional invariants \cite{lax1975periodic, whitham2011linear}. For the truncated system however, $\Lambda < \infty$, the only known invariants are the momentum, energy, and Hamiltonian. The infinite sequence of invariants apart from these first three may no longer be invariants of TKdV.

WRITE PROJECTION OPERATOR AND TKDV

\vsp{10}
%----------------------------------------------------------------------------%
% Hamiltonian
%----------------------------------------------------------------------------%
\subsection{Hamiltonian structure of TKdV}

In examining statistical mechanics of this Hamiltonian system, we will appeal to the idea of a {\em mixed microcanonical-canonical} Gibbs ensemble, as originally introduced in \cite{abramov2003hamiltonian} for the Burgers-Hopf system. Specifically, this ensemble is microcanonical in the energy, with a fixed energy value, and it is canonical in the Hamiltonian. By fixing the energy and hence confining to a compact set, this construction avoids the divergence at infinity that would occur for a simple canonical ensemble. The resulting distribution is thus normalizable.
Accordingly, each Hamiltonian $\Ham^{\pm}$ generates a corresponding mixed ensemble, or Gibbs measure given by 
\begin{align}
\Gupdn = C_{\thupdn} \exp(-\thupdn \Hupdn) \delta(\En - \pi)
\end{align}
Here $\theta = \thupdn$ is the inverse temperature and $C_{\theta}$ a constant (inverse of the partition function) that depends on $\theta$. Each measure has corresponding ensemble mean $\mean{}_{\pm}$.

@@

Recall that the abrupt depth change is met by traveling waves at some time $t = T_{ADC}$, and for simplicity we set $T_{ADC} = 0$. The KdV equations govern the evolution and nonlinear dispersion of waves over long timescales (specifically, $t \gg \delta^{-2}$ for dimensionless time). Meanwhile, wave evolution over shorter timescales, $t \ll \delta^{-2}$, is much simpler in that it is simply noninteracting traveling waves with uniform wave speed $c = \sqrt{g \depth}$ (i.e. no nonlinearity and no dispersion). Hence, a short instant after waves pass over the ADC, the waveform does not yet have time to alter significantly, which gives the (deterministic) matching condition
\begin{align}
&u(x,t) \vert_{t=0^-} = u(x,t) \vert_{t=0^+} \qquad \text{for } x \in [-\pi, \pi]
%&&\mbox{\em deterministic matching condition}
\end{align}
%or in Johnson's formulation, it is $\depth^{1/4} \eta$ that matches.

>>>>

%----------------------------------------------------------%
% Theory
%----------------------------------------------------------%
\section{Theoretical framework}

@@

%----------------------------------------------------------------------------%
% Nondimensionalization
%----------------------------------------------------------------------------%
\subsection{Nondimensionalization and relation to experimental scales}

@@

The evolution of any functional $\Fcnl[u]$ is given by 
\begin{align}
\partial_t \Fcnl = \{ \Fcnl, \Ham \} = \int_{-\pi}^{\pi} \vard{F}{u} \sympJ \vard{\Ham}{u} \dx
\end{align}
where again $\Ham = \Hup$ if $t<0$ and $\Ham = \Hdn$ if $t>0$.

%----------------------------------------------------------------------------%
% Truncated KdV
%----------------------------------------------------------------------------%
\subsection{Truncated KdV}
\label{tKdVSec}

We now introduce the truncated KdV (TKdV) system, which is the main focus of this study. Consider the Fourier series of the state variable
\begin{align}
&u(x,t) = \sum_{k=-\infty}^{\infty} \uhat_k e^{i k x} \, , \\
\label{uhat}
&\uhat_k = \frac{1}{2 \pi} \int_{-\pi}^{\pi} u(x) e^{-i k x} \dx
\end{align}
Since $u$ is real valued, we have $\uhat_{-k} = \uhat_{k}^*$ and since momentum vanishes $\uhat_0 = 0$.
Next, consider the Galerkin truncation at wave number $\Lambda$
\begin{align}
\uL(x,t) = \Proj u =
\sum_{\abs{k} \le \Lambda} \uhat_k e^{i k x} \, , \qquad
\end{align}
where $\Proj$ is a projection operator and \eqref{uhat} still holds. Projecting the KdV equation \eqref{dimlessKdV} onto the finite dimensional space gives the truncated KdV equation (TKdV)
\begin{align}
\label{TKdV}
&\pd{\uL}{t} +  \frac{1}{2} C_3 \drat^{-3/2} \,\Proj (\uL)^2 + C_2 \drat^{1/2} \, \frac{\partial^3 \uL}{\partial x^3} = 0
\qquad \text{for } x \in [-\pi,\pi] \\
&C_3 = \frac{3}{2} \pi^{3/2} \epsup \delup^{-1} \, , \quad
C_2 = \frac{\pi^3 \delup}{6 \lamfac^2} 
\end{align}
Note the additional projection operator in front of the quadratic term $\uL^2$, which removes the aliased modes of wavenumber larger than $\Lambda$. The constants are the same as before and have been repeated here for convenience.

Remarkably, the TKdV system \eqref{TKdV} retains nearly the same Hamiltonian structure described in Section \ref{HamiltonianSection}, with the only modification being the inclusion of the projection operator \cite{bajars2013weakly, majda2019statistical}. The piecewise defined Hamiltonian for TKdV is given by
\begin{equation}
\label{TruncHamiltonian}
\HLupdn = C_2 \dratupdn^{1/2} \, \Htwo[\uL] - C_3 \dratupdn^{-3/2} \, \Hthree[\uL]
\end{equation}
where $\Hthree$ and $\Htwo$ the defined exactly as before \eqref{H3H2}, but now are simply applied to the projected variable $\uL = \Proj u$.
Then TKdV \eqref{TKdV} can be expressed as
\begin{align}
\partial_t {\uL} = \partial_x \Proj \, \vard{\HLupdn}{\uL}
\end{align}
where the truncated symplectic operator is $\SympL = \partial_x \Proj$.
% Note: Bajars 2013 has some important details on the Hamiltonian on the truncated system.

The momentum and energy defined in \eqref{MomEn} remain invariants of TKdV, with the same normalization values $\Mo[\uL] = 0$ and $\En[\uL] = 1$. Note that Parseval's identity gives
\begin{equation}
\En[\uL] = 2 \pi \sum_{k=1}^{\Lambda} \abs{\uhat_k}^2 = 1
\end{equation}
% The above agrees with the first line on top of p. 4 in Majda and Qi J Stat Phys 2019.
As has been noted before, the (untruncated) KdV system possesses an infinite sequence of additional invariants \cite{lax1975periodic, whitham2011linear}. The projected counterparts of these invariants, however, are not necessarily invariants of the truncated system.

% Is it true that the only 3 known invariants of TKdV are the ones mentioned here: momentum, energy, and Hamiltonian.

%----------------------------------------------------------------------------%
% Gibbs
%----------------------------------------------------------------------------%
\subsection{Mixed microcanonical-canonical Gibbs ensemble}

In examining statistical mechanics of this Hamiltonian system, we will appeal to the idea of a {\em mixed microcanonical-canonical} Gibbs ensemble, as originally introduced in \cite{abramov2003hamiltonian} for the Burgers-Hopf system. Specifically, this ensemble is microcanonical in the energy, with a fixed energy value, and it is canonical in the Hamiltonian. By fixing the energy and hence confining to a compact set, this construction avoids the divergence at infinity that would occur for a simple canonical ensemble and hence yields a normalizable distribution. This framework applies equally well to the truncated or untruncated system and hence we will not distinguish between the two.

The Hamiltonian on either side of the ADC, $\Ham^{\pm}$, generates a corresponding mixed ensemble, or Gibbs measure given by 
\begin{align}
\Gupdn = C_{\thupdn} \exp(-\thupdn \Hupdn) \delta(\En - \pi)
\end{align}
Here $\theta = \thupdn$ is the inverse temperature and $C_{\theta}$ a constant (inverse of the partition function) that depends on $\theta$. Each measure has corresponding ensemble mean $\mean{}_{\pm}$.

@@

Recall that the abrupt depth change is met by traveling waves at some time $t = T_{ADC}$, and for simplicity we set $T_{ADC} = 0$. The KdV equations govern the evolution and nonlinear dispersion of waves over long timescales, $t \gg \delta^{-2}$. Meanwhile, wave evolution over shorter timescales, $t \ll \delta^{-2}$, is much simpler in that it is simply noninteracting traveling waves with uniform wave speed $c = \sqrt{g \depth}$ (i.e. no nonlinearity and no dispersion). Hence, a short instant after waves pass over the ADC, the waveform does not yet have time to alter significantly, which gives the (deterministic) matching condition
\begin{align}
&u(x,t) \vert_{t=0^-} = u(x,t) \vert_{t=0^+} \qquad \text{for } x \in [-\pi, \pi]
%&&\mbox{\em deterministic matching condition}
\end{align}
%or in Johnson's formulation, it is $\depth^{1/4} \eta$ that matches.
